Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-05-13 publication_materials/ms_rbc.tex
More edits.

\begin{abstract}
    Natural convection is ubiquitous throughout the physical sciences and engineering, yet many of its important properties remain elusive---particularly in the turbulent regime.
    In this investigation, we derive and solve a quasilinear form of the Rayleigh-Bénard problem by representing the perturbations in terms of marginally stable eigenmodes.
    The amplitude of each eigenmode is determined by requiring that the background state maintains marginal stability.
    The background temperature profile evolves due to the advective flux of each eigenmode, as well as diffusion.
    The entire calculation is one-dimensional, and can be run on a workstation.
    We find the background temperature field evolves to an equilibrium state, where the advective flux from the marginally-stable eigenmodes and the diffusive flux sum to a constant.
    These marginally-stable thermal equilibria (MSTE) are exact solutions of the quasilinear equations.
    The mean temperature profile has thinner boundary layers and larger Nusselt numbers than thermally-equilibrated 2D and 3D simulations of the full nonlinear equations.
    We find the Nusselt number scales like $\Nu \sim\Ra^{1/3}$.
    When an MSTE is used as initial conditions for a 2D simulation, we find that Nu quickly equilibrates without the burst of turbulence often induced by purely conductive initial conditions, but we also find that the kinetic energy is too large and viscously attenuates on a long viscous time-scale.
\end{abstract}

@@

Others have turned to studying quasilinear systems.
The quasilinear approximation starts with a decomposition of all variables into a background and perturbations about this background.
This approximation neglects the influence of nonlinear interactions between the perturbations on the perturbations themselves \cite{marston2016}.
This renders the perturbation equations linear.
Although the quasilinear approximation greatly simplifies the problem, an additional condition must be imposed to determine the amplitude of the perturbations.
In \cite{Beaume_2015}, researchers compute ECS in parallel shear flows by deriving and solving a quasilinear formulation of the Navier-Stokes equations via multi-scale asymptotic arguments. 
They assume the background velocity evolves on a slow time-scale, and to determine the perturbation amplitudes they require marginal stability at each timestep.
A similar strategy is employed by \cite{michel_chini_2019} to studying acoustic streaming.
In that work, an analytic expression for the first-order perturbation's amplitude is found by deriving a solvability condition.

@@

\section{Model Setup}\label{sec:model}
We begin with the Boussinesq approximation for Rayleigh-Bénard Convection, nondimensionalized on the freefall time-scale. 
The domain $\mathcal{D}$ is 2-dimensional, rectangular, and horizontally periodic with spatial dimensions $0 \leq x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two flat boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1/2$ and $-1/2$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\vec{u} = u \hat{x} + w \hat{z} = \vec{0}$ at $z = \pm 1/2$, where $\hat{x}, \hat{z}$ are the unit vectors in the $x$ and $z$ directions. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \vec{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \vec{u}}{\partial t} + \vec{u} \cdot \nabla \vec{u} &= - \nabla p + T \hat{z} + \mathcal{R} \nabla^2 \vec{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \vec{u} \cdot \nabla T &= \mathcal{P} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
For completeness, we specify a final boundary condition $p = p_0$ at $z = - 1/2$. 
Any system of this form can be characterized by its dimensionless Rayleigh number $\Ra = \frac{g\alpha L^3 \Delta T}{\nu \kappa}$ and Prandtl number $\Pr = \frac{\nu}{\kappa}$, where $g, \, \alpha, \, L_z, \, \Delta T, \nu, \kappa$ are the gravitational acceleration, coefficient of thermal expansion, domain height, opposed temperature difference, kinematic viscosity, and thermal diffusivity respectively. 
In this paper, we fix $\Pr = 1$.
For convenience, we define
\begin{equation}
\mathcal{R} = \sqrt{\frac{\Pr}{\Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{\Pr \Ra}}.
\end{equation}

@@

To solve \eq{EQ:T0_IVP} numerically, we need an expression for the perturbation so we can calculate the advective heat flux.
Here we will make the quasilinear approximation, dropping the $\vec{u}'\vec{\cdot}\vec{\nabla}\vec{u}'$ and $\vec{u}'\vec{\cdot}\vec{\nabla}T'$ terms from the evolution equations for the perturbations.
Substituting \eqss{EQ:reynolds_dc_u}{EQ:reynolds_dc_p} into \eqss{EQ:motion1}{EQ:motion3} followed by subtracting \eq{EQ:T0_IVP} and \eq{EQ:p_bar} from the resulting temperature and $\hat{z}$ momentum equations gives
\begin{align}
    \nabla \cdot \vec{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\vec{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \vec{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad u'|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
This is now a linear problem in $\vec{u}'$ and $T'$ which can be solved as an eigenvalue problem.

@@

To calculate the advective heat flux in equation~\ref{EQ:T0_IVP}, we can sum $\langle w'T' \rangle_x$ from each horizontal wavenumber individually.
In this way, the heat flux from the perturbations influence the evolution of $\bar{T}$.
But the evolution of $\bar{T}$ also influences the perturbations, as equation~\ref{EQ:linear3} depends on $\partial_z \bar{T}$.
Thus, the mean temperature and perturbations fields are coupled.

@@

For various $\Ra$ and fixed $\Pr = 1$, we seek marginally-stable thermal equilibria (MSTE) satisfying $\frac{\partial \bar{T}}{\partial t} = 0$ according to \eq{EQ:T0_IVP}. 
We employ the \texttt{Dedalus} pseudo-spectral python framework \cite{Dedalus_2020} to solve the EVP outlined in Section \ref{sec:model} as well as the IVP \eq{EQ:T0_IVP}.
We represent each field with Chebyshev polynomials and use the 3/2 dealiasing rule to calculate the advective heat flux.
The necessary number of basis functions increases with $\Ra$ as the eigenfunctions include increasingly small-scale features. 
We use the \texttt{Eigentools} package \cite{Eigentools} to manipulate the eigenfunctions and calculate the advective heat flux $\langle w' T' \rangle_x$.

At iteration 0 we construct a marginally-stable initial temperature profile $\bar{T}(z, 0)$ whose equation is given in appendix~\ref{sec:initial_profile}. 
At an arbitrary iteration beginning at $t = t_0$, we seek to evolve $\bar{T}(z, t_0)$ into a new marginally-stable profile $\bar{T}(z, t_0 + \Delta t)$ according to \eq{EQ:T0_IVP}. 
To achieve this, we use a second-order, two-stage IMEX Runge-Kutta method.
We initialize \eq{EQ:T0_IVP} with the previous iteration's solution to \eq{EQ:T0_IVP} and maintain marginal stability at each iteration.
The eigenfunctions and amplitude are assumed to be constant over the timestep.
We can solve the IVP over long periods of time by stopping regularly (at each iteration) to adjust the eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $\Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\bar{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing its eigenvalues. 
    Advection tends to stabilize the system, decreasing its eigenvalues. 
    We find the eigenfunction amplitude $A^2$ such that the combination of diffusion and advection yields a new, ``evolved'', marginally-stable mean temperature profile $\bar{T}(z, t_0 + \Delta t)$. In this case $A^2 \approx 1$ due to approximately equal magnitudes of the diffused and advected eigenvalues.}
    \label{fig:iteration_spectra} 
\end{figure}

An iteration is performed as follows.
Consider a marginally-stable temperature profile $\bar{T}(z, t_0)$.
By definition, its maximum eigenvalue is 0.
Diffusing $\bar{T}(z, t_0)$ tends to increase its eigenvalues while ignoring the diffusive term and evolving according to advection tends the stabilize the system.
$A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new IVPs
\begin{align}
    &\frac{\partial \bar{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \bar{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \bar{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = 0 
\end{align}
where $\bar{T}_{\rm{diff}}$ and $\bar{T}_{\rm{adv}}$ denote to the diffused and advected temperature profiles.
We initialize these IVPs with the marginally stable temperature of the current iteration.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then, using the diffused and advected profiles, we compute the eigenvalues $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$.
The appropriate amplitude $A^2$ can then be approximated by 
\begin{equation}
    A^2 \approx A^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.
$\bar{T}(z, t)$ is then evolved according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
Given a fixed timestep $\Delta t$, we assume the dominant eigenvalue can be described by a continuous function $\sigma_{\rm{max}} (A^2)$ which is locally differentiable. 
We use Newton's method to find an amplitude which satisfies our marginal stability tolerance criterion $|\sigma_{\rm{max}} (A^2)| < 10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e. $\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In section \ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.

\subsection{Treatment of Multiple Marginally-Stable Modes} \label{sec:multiple_modes}
In most cases, we encounter eigenvalue spectra with multiple simultaneously marginal modes.
To accommodate this we generalize the advective term in \eq{EQ:T0_IVP} to accommodate $N$ simultaneously marginal modes
\begin{equation}
    \langle w' T' \rangle_x = \sum_{n = 1}^{N} 2 A_n^2  \Re\left[ W_n \theta_n^* \right] 
\end{equation}
where $W_n$ and $\theta_n$ are the eigenmodes associated with $k_x = \frac{n\pi}{2}$. 
There are now $N$ modes, each with their own amplitude to solve for and eigenvalue to keep marginally-stable. 
Given a small fixed time step $\Delta t$, let $\vec{A^2}$ be the amplitude vector and $\vec{\sigma}(\vec{A^2})$ be the dependent eigenvalues.
We expect a function $\vec{\sigma} \, : \, \mathbb{R}^N \to  \mathbb{R}^N$ to have isolated roots $\vec{\tilde{A}^2}$ (should they exist). 
A preliminary approximation $\vec{A^2_0}$ for the appropriate amplitude vector $\vec{\tilde{A}^2}$ is given by a generalized form of \eq{EQ:amp_approx}:
\begin{equation}
    \vec{\tilde{A}^2} \approx \vec{A^2_0} = -\Sigma_{\rm{adv}}^{-1} \vec{\sigma_{\rm{diff}}}.
    \label{EQ:AN_approx}
\end{equation}
Here $\vec{\sigma_{\rm{diff}}} = \vec{\sigma} (\vec{0})$ refers to the eigenvalues after a brief $\Delta t$ period of diffusion. 
We account for the influence of $N$ advection terms on $N$ eigenvalues (one for each mode) by constructing an eigenvalue matrix $\Sigma_{\rm{adv}} \in \mathbb{R}^{N \times N}$.
For example, the element of $\Sigma_{\rm{adv}}$ at row $i$ and column $j$ is given by the $i$th mode's eigenvalue after a $\Delta t$ period of advection by the $j$th mode's eigenfunctions. 
At the point of our preliminary estimate $\vec{A^2_0}$, we approximate the Jacobian matrix 
\begin{equation}
    J = \begin{bmatrix}
        \nabla \sigma_1 (A_1, A_2, ..., A_N) \\
        \nabla \sigma_2 (A_1, A_2, ..., A_N) \\
        \vdots \\
        \nabla \sigma_N (A_1, A_2, ..., A_N) 
    \end{bmatrix}
\end{equation}
via first-order finite differences, requiring $N^2$ more individual EVP solves.
A once-refined estimate $\vec{A^2_1}$ is then given by
% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
\begin{equation}
    \vec{\tilde{A}^2} \approx \vec{A^2_1} = -J^{-1} \vec{\sigma}(\vec{A^2_0}).
\end{equation}
We then use each subsequent guess to adjust the Jacobian via Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the coupling of $\langle w'T' \rangle_x$ with $\bar{T}$.
Over the course of a large time step, $\bar{T}$ evolves according to \eq{EQ:T0_IVP} and eventually the original eigenfunctions cease to provide a stabilizing influence.
It is precisely for this reason why we must halt the IVP regularly to adjust the eigenfunctions and their amplitudes.

Difficulty arises when transitioning between different numbers of marginal modes, particularly when stable modes become marginally-stable mid-iteration.
We facilitate these transitions by defining an adjustable candidate tolerance $\varepsilon_{\rm{cand}} \in [10^{-6}, 10^{-8}]$.
A mode which meets the candidate tolerance can be included as a candidate marginal mode.
Candidate modes are rejected when the root-finding algorithm converges on a negative amplitude $A^2 < 0$.
Otherwise the mode becomes marginally-stable.

After several time steps, $\bar{T}$ tends to asymmetrize due to numerical noise. 
Though asymmetric pairs of solutions may exist, we further constrain $\bar{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each iteration.

@@

Figure~\ref{fig:T0_profiles} gives temperature profiles for $\Ra = 10^8$ where the initial profile, whose construction is outlined in appendix~\ref{sec:initial_profile}, is employed at iteration 0. 
Direct numerical simulations (DNS) are performed by solving \eqss{EQ:motion1}{EQ:motion3} with \texttt{Dedalus}, followed by horizontal- and time- averaging. 
The DNS curve is more diffuse than the MSTE curve, which in turn, is more diffuse than the initial curve. 
Performing an eigenvalue solve by setting $\bar{T}$ equal to the DNS profile yields unstable eigenvalues. 
This suggests that MSTE might maximize boundary layer thickness, subject to the marginal stability constraint.

\begin{figure*}
    \centering
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{flux_sup_n.png}
    \end{tabular}
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{EV_spectra_2ra.png}
    \end{tabular}
    \caption{Heat fluxes (left) and eigenvalue spectra (right) of equilibrated states $\Ra = 2 \times 10^5$ (top) and $\Ra = 10^9$ (bottom). 
    The heat flux profiles are symmetric about $z=0$, so we only plot for $z<0$.
    Advection profiles belong to marginally-stable modes. 
    For low $\Ra$, a single mode with $k_x = 1.5\pi$ is sufficient to oppose boundary layer diffusion and facilitate heat flux throughout the bulk of the domain. 
    For large $\Ra$, high-wavenumber modes contribute pronounced small-scale advection profiles which tightly hug the thin boundary layers. 
    A combination of of progressively wider advection profiles is necessary to transition to the $k_x = 1.5\pi$ mode.}
    \label{fig:flux}
\end{figure*}

@@

Simulations initialized with the MSTE \eq{EQ:mste_ic} do not undergo an early convective transient period, but have faster flows when compared with DNS.
From a dynamical systems perspective, unstable orbits depart from MSTE and approach the global attractor on a viscous time-scale.
This requires more computational effort to achieve relaxation when compared to the conventional conductive initial condition \eq{EQ:linear_ic}.

>>>>

\begin{abstract}
    Natural convection is ubiquitous throughout the physical sciences and engineering, yet many of its important properties remain elusive.
    In this investigation, we derive and solve a quasilinear form of the Rayleigh-Bénard problem by representing the perturbations in terms of marginally stable eigenmodes.
    The amplitude of each eigenmode is determined by requiring that the background state maintains marginal stability.
    The background temperature profile evolves due to the advective flux of each eigenmode, as well as diffusion.
    The entire calculation is one-dimensional, and can be run on a workstation.
    We find the background temperature field evolves to an equilibrium state, where the advective flux from the marginally-stable eigenmodes and the diffusive flux sum to a constant.
    These marginally-stable thermal equilibria (MSTE) are exact solutions of the quasilinear equations.
    The mean temperature profile has thinner boundary layers and larger Nusselt numbers than thermally-equilibrated 2D and 3D simulations of the full nonlinear equations.
    We find the Nusselt number scales like $\Nu \sim\Ra^{1/3}$.
    When an MSTE is used as initial conditions for a 2D simulation, we find that Nu quickly equilibrates without the burst of turbulence often induced by purely conductive initial conditions, but we also find that the kinetic energy is too large and viscously attenuates on a long viscous time-scale.
\end{abstract}

@@

Others have turned to studying quasilinear systems.
The quasilinear approximation starts with a decomposition of all variables into a background and perturbations about this background.
This approximation neglects the nonlinear interactions between perturbations which modify the perturbations themselves \cite{marston2016}.
This renders the perturbation equations linear.
Although the quasilinear approximation greatly simplifies the problem, an additional condition must be imposed to determine the amplitude of the perturbations.
In \cite{Beaume_2015}, researchers compute ECS in parallel shear flows by deriving and solving a quasilinear formulation of the Navier-Stokes equations via multi-scale asymptotic arguments. 
They assume the background velocity evolves on a slow time-scale, and, to determine the perturbation amplitudes, they require marginal stability at each timestep.
A similar strategy is employed by \cite{michel_chini_2019} to study acoustic streaming.
In that work, an analytic expression for the first-order perturbation's amplitude is found by deriving a solvability condition.

@@

\section{Model Setup}\label{sec:model}
We begin with the Boussinesq approximation for Rayleigh-Bénard Convection, nondimensionalized on the freefall time-scale. 
The domain $\mathcal{D}$ is 2-dimensional, rectangular, and horizontally periodic with spatial dimensions $0 \leq x < 4$ and $-1/2 < z < 1/2$. 
The fluid of interest is constrained between two flat boundaries at $z = -1/2$ and $z = 1/2$ with fixed temperatures $1/2$ and $-1/2$ respectively. 
At both boundaries we specify impenetrable, no-slip conditions, such that the velocity $\vec{u} = u \hat{x} + w \hat{z} = \vec{0}$ at $z = \pm 1/2$, where $\hat{x}, \hat{z}$ are the unit vectors in the $x$ and $z$ directions. 
The equations of motion are then given by
\begin{align}
    \nabla \cdot \vec{u} &= 0 \label{EQ:motion1}\\
    \frac{\partial \vec{u}}{\partial t} + \vec{u} \cdot \nabla \vec{u} &= - \nabla p + T \hat{z} + \mathcal{R} \nabla^2 \vec{u} \label{EQ:motion2}\\
    \frac{\partial T}{\partial t} + \vec{u} \cdot \nabla T &= \mathcal{P} \nabla^2 T \label{EQ:motion3}
\end{align}
where $p$ is pressure and $T$ is temperature. 
%For completeness, we specify a final boundary condition $p = p_0$ at $z = - 1/2$. 
Any system of this form can be characterized by its dimensionless Rayleigh number $\Ra = \frac{g\alpha L^3 \Delta T}{\nu \kappa}$ and Prandtl number $\Pr = \frac{\nu}{\kappa}$, where $g, \, \alpha, \, L_z, \, \Delta T, \nu, \kappa$ are the gravitational acceleration, coefficient of thermal expansion, domain height, opposed temperature difference, kinematic viscosity, and thermal diffusivity respectively. 
In this paper, we fix $\Pr = 1$.
For convenience, we define
\begin{equation}
\mathcal{R} = \sqrt{\frac{\Pr}{\Ra}}, \qquad \mathcal{P} = \frac{1}{\sqrt{\Pr \Ra}}.
\end{equation}

@@

To solve \eq{EQ:T0_IVP} numerically, we need an expression for the perturbations so we can calculate the advective heat flux.
Here we will make the quasilinear approximation, dropping the $\vec{u}'\vec{\cdot}\vec{\nabla}\vec{u}'$ and $\vec{u}'\vec{\cdot}\vec{\nabla}T'$ terms from the evolution equations for the perturbations.
Substituting \eqss{EQ:reynolds_dc_u}{EQ:reynolds_dc_p} into \eqss{EQ:motion1}{EQ:motion3} followed by subtracting \eq{EQ:T0_IVP} and \eq{EQ:p_bar} from the resulting temperature and $\hat{z}$ momentum equations gives
\begin{align}
    \nabla \cdot \vec{u'} &= 0 \label{EQ:linear1}\\
    \frac{\partial\vec{u'}}{\partial t} &= - \nabla p' + T'\hat{z} + \mathcal{R} \nabla^2 \vec{u'} \label{EQ:linear2}\\
    \frac{\partial T'}{\partial t} + \frac{\partial \bar{T}}{\partial z} w' &= \mathcal{P} \nabla^2 T' \label{EQ:linear3}
\end{align}
with Dirichlet boundary conditions 
\begin{equation}
    T'|_{z = \pm \frac{1}{2}} = 0, \quad u'|_{z = \pm \frac{1}{2}} = 0, \quad p'|_{z = \pm \frac{1}{2}} = 0.
\end{equation}
This is now a linear problem in $\vec{u}'$ and $T'$ which can be solved as an eigenvalue problem.

@@

To calculate the advective heat flux in equation~\ref{EQ:T0_IVP}, we can sum $\langle w'T' \rangle_x$ from each horizontal wavenumber individually.
In this way, the heat flux from the perturbations influence the evolution of $\bar{T}$.
But the evolution of $\bar{T}$ also influences the perturbations, as equation~\ref{EQ:linear3} depends on $\partial_z \bar{T}$.
Thus, the mean temperature and perturbation fields are coupled.

@@

For various $\Ra$ and fixed $\Pr = 1$, we seek marginally-stable thermal equilibria (MSTE) satisfying $\frac{\partial \bar{T}}{\partial t} = 0$ according to \eq{EQ:T0_IVP}. 
We employ the \texttt{Dedalus} pseudo-spectral python framework \cite{Dedalus_2020} to solve the eigenvalue problem outlined in Section \ref{sec:model} as well as the time evolution equations \eq{EQ:T0_IVP}.
We represent each field with Chebyshev polynomials and use the 3/2 dealiasing rule to calculate the advective heat flux.
The necessary number of basis functions increases with $\Ra$ as the eigenfunctions include increasingly small-scale features. 
We use the \texttt{Eigentools} package \cite{Eigentools} to manipulate the eigenfunctions and calculate the advective heat flux $\langle w' T' \rangle_x$.

To initialize our calculations, we construct a marginally-stable initial temperature profile $\bar{T}(z, t=0)$ whose equation is given in appendix~\ref{sec:initial_profile}. 
We use \eq{EQ:T0_IVP} to evolve $\bar{T}(z, t_0)$ into a new marginally-stable profile $\bar{T}(z, t_0 + \Delta t)$.
We use a second-order, two-stage IMEX Runge-Kutta method.
The eigenfunctions and amplitudes are assumed to be constant over the timestep.
We symmetrize $\bar{T}$ by setting the coefficients of its even Chebyshev basis functions to zero at the start of each timestep.
This ensures $\bar{T}(z) = -\bar{T}(-z)$.
On every timestep, we calculate the marginally stable eigenfunctions and their amplitude $A$.
It is essential that we pick the correct eigenfunction amplitude when calculating the advective term to maintain marginal stability.
We will now illustrate our method of finding the appropriate $A$ through an example.

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $\Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\bar{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing $\sigma$. 
    Advection tends to stabilize the system, decreasing $\sigma$. 
    We find the eigenfunction amplitude $A^2$ such that the combination of diffusion and advection yields a new, ``evolved,'' marginally-stable mean temperature profile $\bar{T}(z, t_0 + \Delta t)$. In this case $A^2 \approx 1$ due to approximately equal magnitudes of the diffused and advected eigenvalues.}
    \label{fig:iteration_spectra} 
\end{figure}

Consider a marginally-stable temperature profile $\bar{T}(z, t_0)$.
By definition, its maximum growth rate is 0.
Diffusing $\bar{T}(z, t_0)$ tends to increase its growth rates while ignoring the diffusive term and evolving according to advection tends to decrease its growth rates.
$A$ must be selected such that these two influences are equal and opposite.
We can measure the effects of diffusion and advection on the maximum eigenvalue by solving two new IVPs
\begin{align}\label{eqn:diffusion_only}
    &\frac{\partial \bar{T}_{\rm{diff}}}{\partial t} = \mathcal{P} \frac{\partial^2 \bar{T}_{\rm{diff}}}{\partial z^2} \\
    &\frac{\partial \bar{T}_{\rm{adv}}}{\partial t} + \frac{\partial}{\partial z} \langle w'T' \rangle_x = 0 
\end{align}
where $\bar{T}_{\rm{diff}}$ and $\bar{T}_{\rm{adv}}$ denote to the diffused and advected temperature profiles.
We initialize $\bar{T}_{\rm{diff}}$ and $\bar{T}_{\rm{adv}}$ with $\bar{T}(z, t_0)$, and evolve the equations to $t=t_0+\Delta t$.
Suppose $k_0$ is the wavenumber of the marginally-stable mode.
Then, using the diffused and advected profiles, we compute the eigenvalues $\sigma_{\rm{diff}}$ and $\sigma_{\rm{adv}}$ at $k_x = k_0$ and $t=t_0+\Delta t$.
The appropriate amplitude $A^2$ can then be approximated by 
\begin{equation}
    A^2 \approx A^2_0 \equiv -\frac{\sigma_{\rm{diff}}}{\sigma_{\rm{adv}}} \label{EQ:amp_approx}
\end{equation}
where the square in $A^2_0$ is due to the quadratic nonlinearity of the eigenfunctions in the advective term and the minus sign is due to the fact that $\sigma_{\rm{adv}} < 0$ and $\sigma_{\rm{diff}} > 0$ in general.
We illustrate these trends in Figure~\ref{fig:iteration_spectra}. 
The subscript 0 denotes a preliminary estimate which we later refine.

Using the amplitude $A_0^2$, the background temperature $\bar{T}(z, t_0)$ is then evolved to $t=t_0+\Delta t$ according to \eq{EQ:T0_IVP} and another eigenvalue solve is performed. 
This background temperature is typically close to, but not exactly, marginally stable.
We use Newton's method to find an amplitude $A^2$ such that the maximum growth rate is zero to within $10^{-9}$.
Marginally-stable modes do not oscillate in time, i.e. $\sigma = 0$ implies $\omega = 0$.
This agrees with the conventional notion of exchange of stabilities \cite{drazin_reid_2004}.
Crucially, we do not assume the $k_x$ of the marginally-stable mode is fixed.
In section \ref{sec:multiple_modes} we specify procedures for the treatment of multiple simultaneously marginal modes.

\subsection{Treatment of Multiple Marginally-Stable Modes} \label{sec:multiple_modes}
In most cases, we encounter eigenvalue spectra with multiple marginal modes.
To accommodate this we generalize the advective term in \eq{EQ:T0_IVP} to accommodate $N$ simultaneously marginal modes
\begin{equation}
    \langle w' T' \rangle_x = \sum_{n = 1}^{N} 2 A_n^2  \Re\left[ W_n \theta_n^* \right] 
\end{equation}
where $W_n$ and $\theta_n$ are the marginally stable eigenmodes associated with different values of $k_x$. 
There are now $N$ modes, each with their own amplitude to solve for and eigenvalue to keep marginally-stable. 
Given a small fixed time step $\Delta t$, let $\vec{A^2}$ be the amplitude vector and $\vec{\sigma}(\vec{A^2})$ be the eigenvalues of the $N$ modes.
We are searching for $\vec{\tilde{A}^2}$ that satisfies of $\vec{\sigma}(\vec{\tilde{A}^2})=0$.
We approximate the amplitudes with $\vec{A^2_0}$ by generalizing \eq{EQ:amp_approx}:
\begin{equation}
    \vec{\tilde{A}^2} \approx \vec{A^2_0} = -\Sigma_{\rm{adv}}^{-1} \vec{\sigma_{\rm{diff}}}.
    \label{EQ:AN_approx}
\end{equation}
Here $\vec{\sigma_{\rm{diff}}} = \vec{\sigma} (\vec{0})$ refers to the eigenvalues of the background temperature after evolving under only diffusion (equation \ref{eqn:diffusion_only}) for a time interval $\Delta t$.
We account for the influence of $N$ advection terms on $N$ eigenvalues (one for each mode) by constructing an eigenvalue matrix $\Sigma_{\rm{adv}}$.
To calculate element of $\Sigma_{\rm{adv}}$ at row $i$ and column $j$, we evolve the background temperature under only advection by mode $j$ for a time interval $\Delta t$.
Then the $i,j$ component of $\Sigma_{\rm{adv}}$ is given by the eigenvalue of the $i$th mode.

Our estimate $\vec{A^2_0}$ does not give a background temperature for which all $N$ modes are exactly marginally stable.
We refine our vector of amplitudes via Newton's method.
This requires calculating the Jacobian matrix 
\begin{equation}
    J = \begin{bmatrix}
        \nabla \sigma_1 (A_1, A_2, ..., A_N) \\
        \nabla \sigma_2 (A_1, A_2, ..., A_N) \\
        \vdots \\
        \nabla \sigma_N (A_1, A_2, ..., A_N) 
    \end{bmatrix}.
\end{equation}
We approximate $J$ via first-order finite differences, which requires an additional $N^2$ sparse eigenvalue solves.
%A once-refined estimate $\vec{A^2_1}$ is then given by
%% The $k$th row of $J_0$ is then given by the gradient of the $k$th mode's eigenvalue $\nabla \sigma_k$.
%\begin{equation}
%    \vec{\tilde{A}^2} \approx \vec{A^2_1} = -J^{-1} \vec{\sigma}(\vec{A^2_0}).
%\end{equation}
%We then use each subsequent guess to adjust the Jacobian via Broyden's method for root-finding in multi-dimensional functions \cite{Broyden}.
We iterate Newton's method until all marginally stable modes have eigenvalues within $10^{-9}$ of zero.
We find $\vec{\sigma}(\vec{A^2})$ does indeed have a unique root provided the time step is not too large and there are no numerical instabilities, as outlined in appendix \ref{sec:timestep}.
Presumably this is due to the coupling of $\langle w'T' \rangle_x$ with $\bar{T}$.
Over the course of a large time step, $\bar{T}$ evolves according to \eq{EQ:T0_IVP} and eventually the original eigenfunctions cease to provide a stabilizing influence.
This limits the timestep size we can take.

Difficulty arises when transitioning between different numbers of marginal modes.
We facilitate these transitions by defining an adjustable candidate tolerance $\varepsilon_{\rm{cand}} \in [10^{-6}, 10^{-8}]$.
A mode which meets the candidate tolerance can be included as a candidate marginal mode.
Candidate modes are rejected when the root-finding algorithm converges on a negative amplitude $A^2 < 0$.
Otherwise the mode becomes marginally-stable.

@@

Figure~\ref{fig:T0_profiles} gives temperature profiles for $\Ra = 10^8$, where the initial profile is outlined in appendix~\ref{sec:initial_profile}. 
We ran a direct numerical simulation (DNS) of \eqss{EQ:motion1}{EQ:motion3} with \texttt{Dedalus}, and also plot the horizontal- and time- averaged temperature profile. 
The temperature boundary layers in the DNS are wider than in the MSTE, which is in turn wider than the boundary layer of the initial temperature profile.
Performing an eigenvalue solve by setting $\bar{T}$ equal to the DNS profile yields unstable eigenvalues. 
This suggests that MSTE might maximize boundary layer thickness, subject to the marginal stability constraint.

\begin{figure*}
    \centering
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{flux_sup_n.png}
    \end{tabular}
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{EV_spectra_2ra.png}
    \end{tabular}
    \caption{Heat fluxes (left) and eigenvalue spectra (right) of equilibrated states with $\Ra = 2 \times 10^5$ (top) and $\Ra = 10^9$ (bottom). 
    The heat flux profiles are symmetric about $z=0$, so we only plot for $z<0$.
    We plot the advection profiles for marginally-stable modes. 
    At low $\Ra$, a single mode with $k_x = 1.5\pi$ is sufficient to oppose boundary layer diffusion and facilitate heat flux throughout the bulk of the domain. 
    At large $\Ra$, high-wavenumber modes contribute pronounced small-scale advection profiles which tightly hug the thin boundary layers. 
    A combination of of progressively wider advection profiles is necessary to transition to bulk advection provided by the $k_x = 1.5\pi$ mode.}
    \label{fig:flux}
\end{figure*}

@@

Simulations initialized with the MSTE \eq{EQ:mste_ic} do not undergo an early convective transient period, but have faster flows when compared with DNS.
From a dynamical systems perspective, unstable orbits depart from MSTE and approach the global attractor on a viscous time-scale.
This requires more computational effort to achieve relaxation when compared to the conventional conductive initial condition \eq{EQ:linear_ic}.
Initializing with only the temperature state of the MSTE avoids an early convective transient, but does not have excessive large-scale flows.
These state could be excellent initial conditions for high-Rayleigh number DNS.
